Extracting sunbeams from cucumbers: Computational reproducibility is essential for archaeological science

### 2025-02-03 paper.qmd
edits

---
title: "Extracting sunbeams from cucumbers: Reproducibility as a necessity for the future of archaeological science"
format: html
bibliography: references.bib
fig-dpi: 600
execute: 
    echo: false
    warning: false
---

@@

In their paper celebrating the 40th anniversary of this journal Torrence et al [-@torrenceFortyYearsStill2015] noted that reproducibility was an issue important to the reputation and sustainability of the discipline, and necessary for archaeological science to behave like a science. As part of the celebration of the 50th anniversary, and of Torrence's leadership of the journal, my contribution revisits these topics of archaeology's status as a science, this journal's place in the landscape of archaeological science, and how the journal has responded to a growing recognition of the importance of reproducibility. I first present bibliometric evidence of the position of archaeology as a whole, and this journal in particular, in the hierarchy of the sciences. Next I report on the journal's progress in supporting reproducible research, and my work doing a new kind of peer review for JAS, one that evaluates the computational reproducibilty of the research submitted for publication.

@@

At the time of writing (Feburary 2025) I have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS. Of these, 11 have been published in JAS to date. Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate, by a generous measure, of 63%. Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors. For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate. Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), which are articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g. instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%. In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values obtained from a variety of harder and soft sciences.

@@

>>>>

---
title: "Extracting sunbeams from cucumbers: Computational reproducibility is essential for archaeological science"
format: html
bibliography: references.bib
fig-dpi: 600
execute: 
    echo: false
    warning: false
---

@@

In their paper celebrating the 40th anniversary of this journal Torrence et al [-@torrenceFortyYearsStill2015] noted that reproducibility was an issue important to the reputation and sustainability of the discipline, and necessary for archaeological science to behave like a science. As part of the celebration of the 50th anniversary, and of Torrence's leadership of the journal, my contribution revisits these topics of archaeology's status as a science, this journal's place in the landscape of archaeological science, and how the journal has responded to a growing recognition of the importance of reproducibility. I first present bibliometric evidence of the position of archaeology as a whole, and this journal in particular, in the sciences. Next I report on the journal's progress in supporting reproducible research, and my work doing a new kind of peer review for JAS, one that evaluates the computational reproducibilty of the research submitted for publication. Finally, I analyse twelve months of reproducibility reviews to identify common weaknesses in the ways archaeologists are working currently, and provide simple recommendations for researchers to overcome these and contribute to the improvement computational reproducibility in archaeological science.

@@

At the time of writing (Feburary 2025) I have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS. Of these, 11 have been published in JAS to date. Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate, by a generous measure, of 63%. Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors. For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate. Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), the success rate is 7%. In my opinion this is the proper denominator because these are articles that report quantitative analyses, and all quantitative analyses should be done using an open source programming language to provide transparency (those authors used commercial software such as Microsoft Excel or SPSS, etc.). In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values obtained from a variety of harder and soft sciences.

@@

While consensus is one measure of the hardness of a science, the computability
